---
title: "NewMarkdown"
author: "Benjamin Zeiger"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
```

Markdown files always retain the folder they are saved in as the default working directory

```{r}
getwd()
```

So, we dont need to specify a folder if we want to access a document adjacent to the .Rmd file.

```{r}

read_excel("TestSheet.xlsx")

```


We can change the directory manually, but it only lasts for one code chunk:

```{r}

setwd("C:/Users/zeige/Documents/School/ENR_PhD_Courses/6193_Stat/git/ENR6193_Biostats/TestFolder/data_files")

read_excel("data_excel.xlsx")

```


Note above that RStudio warns us the directory will be *reset after the chunk finishes running*.

So, if we want to access the "data_excel" document, we need to specify the folder again.

```{r}

read_excel("data_excel.xlsx")

```

But we can still access the TestSheet doc without specifying path

```{r}

read_excel("TestSheet.xlsx")

```




