---
title: "GLMM Mini-Workshop: Tick Movement Assay"
author: "Troy Koser"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE) #this is just telling r markdown to please not write out warning or error messages when knitting to html or pdf
```

# Regression philosophy (Regression and Other Stories Chapters 6-8)

So, what is regression and why do we hear so much about it nowadays?
Part of the rise of regression from my view has to do with the suite of
packages in R that mostly handle regression analyses (glmmTMB, lme4,
some others I'm forgetting) and their ability to handle multiple types
of data. Many classical statistical tests which relied on normal
distribution assumptions can be expressed as special forms of a
regression, so instead of running an ANOVA in this package and a linear
model in that one you could use a single package to run everything you
might need.

A classic regression analysis is set up to answer the question "How does
the expected value of a response change as a function of the given
predictors" and follows the form below for linear regression. $$
Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip} + \varepsilon_i
$$

-   $Y_i$: the response (outcome) value for observation $i$
-   $X_{ij}$: the value of predictor $j$ for observation $i$ (e.g.,
    temperature, treatment indicators)
-   $p$: the number of predictors in the model
-   $\beta_0$: the intercept (expected value of $Y$ when all predictors
    are 0 / at their reference levels)
-   $\beta_j$: the coefficient for predictor $j$ (expected change in $Y$
    for a 1-unit increase in $X_{ij}$, holding other predictors
    constant)
-   $\varepsilon_i$: the residual (unexplained deviation for observation
    $i$), typically assumed $\varepsilon_i \sim \mathcal{N}(0,\sigma^2)$
-   $\sigma^2$: the residual variance (how spread out the residuals are
    around 0)

One of the most important things to understand about regression modeling
is the residual variance. The graphs below will demonstrate what
residuals are and how we look at their distributions.

```{r echo=FALSE}
###############################################################
# 1) Linear model with Gaussian (normal) errors
###############################################################

library(tidyverse)
library(MASS)   # helps us make the negbin distribution/model

set.seed(123)

# Simulate predictor
x <- seq(0, 10, length.out = 50)

# True relationship
beta0 <- 2
beta1 <- 1.5

# Gaussian noise
epsilon <- rnorm(50, mean = 0, sd = 2)

# Response
y <- beta0 + beta1 * x + epsilon

dat_lm <- tibble(x = x, y = y)

# Fit linear model
m_lm <- lm(y ~ x, data = dat_lm)

# Add fitted values and residuals
dat_lm <- dat_lm %>%
  mutate(
    fitted = predict(m_lm),
    residual = y - fitted
  )

###############################################################
# Plot linear model with residuals labeled
###############################################################

ggplot(dat_lm, aes(x, y)) +
  geom_point(size = 2) +
  geom_line(aes(y = fitted), color = "blue", size = 1) +
  geom_segment(aes(x = x, xend = x,
                   y = fitted, yend = y),
               linetype = "dashed", alpha = 0.6) +
  annotate("text", x = 8, y = 15,
           label = "Vertical distance = Residual (Error)",
           size = 4) +
  theme_bw() +
  labs(title = "Linear Model with Gaussian Errors",
       y = "Response",
       x = "Predictor")
```

You can see in this graph that most of our points are around our fit
line with some variation in how far they are from the line. That
variation in how far our data points are from our linear model
predictions is called the residuals or sometimes the error! We can graph
what this looks like using a frequency histogram, which will let us know
if our model is "wrong" in ways that are somewhat consistent/normal.

```{r echo=FALSE}
###############################################################
# 2) Visualize normal error distribution
###############################################################

errors <- rnorm(10000, mean = 0, sd = 2)

ggplot(data.frame(errors), aes(errors)) +
  geom_histogram(aes(y = ..density..), bins = 40) +
  stat_function(fun = dnorm,
                args = list(mean = 0, sd = 2),
                size = 1) +
  theme_bw() +
  labs(title = "Normal Error Distribution",
       x = "Error",
       y = "Density")
```

Soooooooo our data like NEVER look like this!!!! The generalized linear
model is cool and everything but our data points are usually varying in
how far they are away from our model predictions in funky ways that mean
we get to use generalized linear models! An example can be seen below
with a negative binomial count dataset and its error structure.

```{r echo=FALSE}
###############################################################
# 3) Linear predictor with Negative Binomial error structure
###############################################################

# Simulate predictor
x_nb <- seq(0, 10, length.out = 50)

# True log-scale relationship (for counts)
beta0_nb <- 1
beta1_nb <- 0.3

# Expected mean (log link)
mu <- exp(beta0_nb + beta1_nb * x_nb)

# Generate negative binomial counts
y_nb <- rnegbin(50, mu = mu, theta = 1.5)

dat_nb <- tibble(x = x_nb, y = y_nb)

# Fit negative binomial model
m_nb <- glm.nb(y ~ x, data = dat_nb)

# Add predicted mean
dat_nb <- dat_nb %>%
  mutate(
    fitted = predict(m_nb, type = "response")
  )

###############################################################
# Plot Negative Binomial model
###############################################################

ggplot(dat_nb, aes(x, y)) +
  geom_point(size = 2) +
  geom_line(aes(y = fitted), color = "red", size = 1) +
  theme_bw() +
  labs(title = "Negative Binomial GLM (Count Data)",
       y = "Count Response",
       x = "Predictor")

###############################################################
# Optional: Visualize negative binomial distribution shape
###############################################################

x_vals <- 0:20
nb_density <- dnbinom(x_vals, size = 1.5, mu = 5)

ggplot(data.frame(x_vals, nb_density),
       aes(x_vals, nb_density)) +
  geom_col() +
  theme_bw() +
  labs(title = "Negative Binomial Distribution",
       x = "Count",
       y = "Probability")

```

Typically linear models would be used specifically for analyzing the
relationship between a continuous predictor (like temperature) and a
continuous response (like height). In classic linear regression, the
error/residuals are assumed to be approximated by the normal
distribution which may or may not be a valid assumption. "Generalizing"
the linear model means we move beyond assuming a normal distribution for
our response and error distributions and instead consider different
distributions like Guassian, binomial, Poisson, negative binomial, etc.

The GLM extends on the linear model in two important ways: by allowing
for different error distributions and by allowing for different "link
functions" which describe how predictors map to the mean. Common link
functions include:

**Identity (Linear)**

**Log (Multiplicative effect)**

**Logit (Probabilities between 0 and 1)**

Unlike many stats classes, we won't pretend there will be a perfect fit
of analysis to data type here, instead we'll use the workflow below to
explore how we can be constantly improving our analysis approach and
looking for better ways to handle our data.

The two most common packages for running generalized linear models
(GLMs) are lme4 and glmmTMB. lme4 is the classic, the OG, the workhorse,
and it is widely used and well-documented. It fits both random
intercepts and slopes for mixed-effect models (we'll talk about this
later) and has functionality for fitting Gaussian, Poisson, and binomial
error distributions. glmmTMB is the young upstart, fast and more
unstable (honestly same) and a bit less well-documented but still quite
common in ecology papers nowadays. glmmTMB allows for more zero-inflated
error distribution fitting options (Nbinom, hurdle, zeroinflated, etc.)
and allows for some cool structuring in your random effects (more than
just nested). glmmTMB also notoriously will NOT yell at you about
overfitting your model (more predictors/more complex structure than you
have data/variance to analyze with). You can probably tell that I'm a
glmmTMB gurly at heart, I'll probably use it whenever we are doing GLMMs
but lmk if you want some more lme4 worked examples.

In this exercise, we’re going to analyze data from a controlled tick
movement arena assay. This is deliberately a little toy-ish, but it’s
structured to explore the kinds of issues you actually run into with
real ecological data plus it might be useful for some of Liz's work!

# Experiment simulation/exercise setup

Here’s the setup we’re pretending happened in the lab:

\-**Each trial releases 15 adult ticks into a circular arena.**

**One of four repellents (or a control with no repellent) is applied to
a specific zone of the arena.**

**Over a 2-minute observation period, each tick is classified as
either:**

**1 = spent the majority of the time in the half of the arena near the
repellent**

**0 = spent the majority of the time in the half of the arena away from
the repellent**

This is a good example of slight data manipulation at the experimental
level. We might have super fine-scale data on tick movement, speed,
direction, etc. but at some point we need to analyze this stuff so we
often simplify. Some common simplification reasons include:

**Matching response variable to scale to predictor scales (ex. tick
counts as GPS points but predictors are averaged at county level)**

**Mismatch between scientific research questions and granularity of
response data (do I care about the nuances/variability of tick or mite
movement behavior or do I care about repellent action?)**

**INTERPRETATION!!!!! Are you planning on talking about your results to
managers or having other scientists look/understand your approach? If
so, making your analysis simple and easy to interpret is an important
consideration!**

**Ran out of time/had to complete the PhD**

A note of simplifying your data: you can always make granular data more
simple (I estimated hairloss on a scale of 0-100 but categorized values
on an index from 1-5 for my phd), but you usually cannot make simple
data more granular. If you're messing around with response data always
remember to save your raw data separately in its own forbidden/no-touchy
folder for your project. Common simplifications include the binary move
we made in this example, clustering year/age values in age classes
(juvenile, subadult, adult), or other thresholds.

So... tangent aside: our response is binary at the tick level, but with
strong grouping:

**ticks are nested within trials**

**trials are run in 5 physical arenas**

Importantly, we’re going to bake in an arena effect on purpose. Some
arenas are just... weird. Think about that one tick-y, fly-y, or
chigger-y site where you know you'll find a horde of individuals but you
can't explain why. For this arena example, maybe the lighting is off,
maybe airflow is different, maybe the floor texture is subtly different.
Whatever the reason, ticks in some arenas are more likely to drift
toward the repellent zone regardless of treatment. That’s not a
nuisance, that’s ecological structure, baby!

We also include a continuous fixed effect, temperature (°C), because
basically every biological process gets dragged around by temperature,
and it gives us a clean example of how continuous predictors slot into
GLMs and GLMMs. Doesn't make the most sense in our example but it'll be
a good modeling addition to talk about.

So before we fit anything, we’re already making modeling decisions:

**binary outcome → binomial error structure**

**nested data → mixed effects**

**known sources of heterogeneity → random effects**

*NOW YOU TRY (FRANTIC TEACHER NOISES)!*

In 2–3 sentences: In Regression and other Stories the authors are
constantly talking about regressions as "generative models"...wtf does
that mean?

Your response: If we have well-informed predictors (i.e. the process is
"known") we can use the regression to "generate" a new dataset. So if we
can accurately predict insect flight activity using temperature and
r.h., then we can draw a new "sample" of activities based on given
environmental records.

##2.1 Simulate an example dataset (tick-level binary outcomes)\##

```{r}
library(lme4)
library(glmmTMB)
library(broom.mixed)
```

```{r}
set.seed(123)

# Five physical arenas
arenas <- c("Arena1", "Arena2", "Arena3", "Arena4", "Arena5")

trial_plan <- tibble(
  treatment = c("Control", "RepA", "RepB", "RepC", "RepD"),
  n_trials  = c(3, 6, 6, 6, 6)
)

trials <- trial_plan %>%
  uncount(n_trials, .id = "trial_num") %>%
  mutate(
    trial_id = paste0(treatment, "_", trial_num),
    arena    = sample(arenas, n(), replace = TRUE),
    temp_c   = runif(n(), 18, 28)
  )

# Arena-specific bias toward the repellent zone
# (Arena2 and Arena4 are a little cursed)
arena_effect <- c(
  Arena1 =  0.0,
  Arena2 =  0.9,
  Arena3 = -0.3,
  Arena4 =  0.6,
  Arena5 = -0.2
)

# Continuous temperature effect
beta_temp <- 0.12

# Treatment effects (relative to control) on the logit scale <-why logit here?
beta_trt <- c(
  Control =  0.0,
  RepA    = -0.7,
  RepB    = -1.2,
  RepC    = -0.4,
  RepD    = -0.9
)

# Baseline probability ~0.55 in Control, average arena, ~23C
intercept <- qlogis(0.55)

tick_dat <- trials %>%
  slice(rep(1:n(), each = 15)) %>%
  group_by(trial_id) %>%
  mutate(tick_id = row_number()) %>%
  ungroup() %>%
  mutate(
    eta = intercept +
      beta_trt[treatment] +
      beta_temp * (temp_c - 23) +
      arena_effect[arena],
    p_toward = plogis(eta),
    toward   = rbinom(n(), size = 1, prob = p_toward)
  )

glimpse(tick_dat)

```

TIME FOR YOU TO CODE!

Write code to answer: 1) How many trials are there?\
2) How many ticks total?\
3) Confirm that each trial has 15 ticks (show a small table).

```{r}
# YOUR CODE HERE

tick_conf <- tick_dat %>%
  group_by(trial_id) %>%
  summarize(ticks_per = n_distinct(tick_id))

tick_conf

tick_conf %>% summarize(trials = n_distinct(trial_id),
                       tick_tot = sum(ticks_per))

```

# Exploratory data analysis (EDA): learning what the data want to be

Before we touch a model, we need to spend some time with the data
themselves. EDA is where we start figuring out:

**What the observational unit really is**

**What kinds of variation exist in the data**

**Whether the structure we think we imposed experimentally actually
shows up**

Remember from lab 2: we aren't fixing problems, we are molding our model
around the bumps and grooves of our dataset. In a perfect world you
would simulate your data and effect sizes, design a sampling regime or
experiment to robustly test your expected relationships, then simply run
your analysis on your collected data like you did with your simulated
data. This is a fast-growing idea in disease ecology called
"model-guided field work"
(<https://doi.org/10.1111/j.1461-0248.2012.01836.x>,
<https://doi.org/10.1890/070086>). Most of the time we aren't so
lucky...we end up having very little information (or, in the case of
grad students, limited skill sets in the early years) to accurately
model the relationships we'd like to explore. We end up going out and
collecting some data with a research question in mind then show up in
front of R hoping we can answer some questions with the data we have. In
this course we'll largely stick to this idea that we are exploring data
and looking for hints as to the best ways to analyze given its structure
and our questions, but know that there exists an ideal scenario where
you may feel comfortable enough with R and know enough about your system
to simulate data -\> design experiments -\> analyze -\> repeat. If this
sounds like a utopia you'd like to explore, lmk and we can chat...

In our case, we simulated tick-level binary outcomes, but the biology
and experimental design introduce structure at higher levels (trial,
arena). EDA helps us decide whether those structures are likely to
matter.

```{r}
trial_sum <- tick_dat %>%
  group_by(trial_id, treatment, arena, temp_c) %>%
  summarise(
    n_ticks = n(),
    n_toward = sum(toward),
    prop_toward = mean(toward),
    .groups = "drop"
  )

head(trial_sum)

```

## 3.1 Outcome balance

A very early (but very important) question for binary data is: Do we
have a reasonable mix of 0s and 1s?

```{r}
tick_dat %>%
  count(toward) %>%
  mutate(prop = n / sum(n))

```

Makes sense, right? We are testing repellents, it makes sense that the
majority of ticks would be moving away from our repellents.

*Your turn!* Is the outcome roughly balanced or highly skewed? Why might
imbalance matter in logistic regression?

Your response:

## 3.2 Treatment × arena patterns

Now we start looking for structure we know exists, because we built it
into the simulation on purpose.

```{r}
trial_sum %>%
  ggplot(aes(treatment, prop_toward)) +
  geom_point(aes(shape = arena),
             position = position_jitter(width = 0.1),
             alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  theme_bw()
```

**Figure interpretation**

Which repellent appears most effective?

Is there evidence of arena-level bias?

Your response: Repellent B seems most effective from this view. Arena 2
seems to have a consistently higher proportion of "towards"
observations.

This plot is already hinting at something important: treatment effects
are not the only source of variation here.

## 3.3 Arena effects by treatment

Let’s look at arenas more directly.

```{r}
trial_sum %>%
  ggplot(aes(arena, prop_toward)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.7) +
  facet_wrap(~treatment) +
  theme_bw()

```

**Interpretation**

Which arena appears systematically different?

Your response: Arena 2 has a higher average prop_toward than others.

At this point, it should be pretty clear that arena matters — but we
probably don’t care about Arena 2 as a specific, special snowflake. We
care that arenas as a class introduce variability.

That distinction is the entire motivation for random effects! We could
create separate models for each arena, but we would lose some overall
power in our ability to answer our research question on repellent
activity. Adding arena as a random effect, in other words a grouping
variable that impacts our response variable in a way we don't explicitly
want to examine for its impact (I don't care which arena had the most 0
vs 1 movement responses), will help account for this extra between-group
variance!

## 3.4 Temperature as a continuous effect

We also included temperature as a continuous predictor, because real
ecological systems almost always respond to gradients.

```{r}
trial_sum %>%
  ggplot(aes(temp_c, prop_toward)) +
  geom_point(aes(shape = arena)) +
  geom_smooth(method = "lm") +
  facet_wrap(~treatment) +
  theme_bw()

```

**Interpretation**

Is temperature plausibly related to tick movement?

Your response: There appear to be consistent positive relationships
between temp and movement towards repellent.

At this stage we are not deciding on a final model — we’re asking
whether temperature is even worth carrying forward as a predictor. What
might be a good reason why we would carry temperature forward even if we
did not see a strong signal at this stage?

Response: It is a simple, well-known predictor of arthropod activity,
and the model may be able to see an interaction between temp and
repellent that we can't quit parse with the naked eye.

# Choosing distributions and link functions

Once we understand the data structure, we can start thinking about how
to model it.

Our response variable is:

**binary (toward / away)**

**measured at the tick level**

That immediately suggests a binomial likelihood.

Write code to confirm that: 1) `toward` is binary (only 0 and 1)\
2) `prop_toward` is a proportion bounded [0,1]\
3) `n_toward` is an integer count from 0 to 15

```{r}
# YOUR CODE HERE

tick_dat %>% distinct(toward)

trial_sum %>% summarise(propMax = max(prop_toward),
                        propMin = min(prop_toward),
                        propTrue = is.double(prop_toward))

trial_sum %>% summarise(towardMax = max(n_toward),
                        towardMin = min(n_toward),
                        intTrue = is.integer(n_toward))


```

But GLMs require two choices:

**an error distribution**

**a link function**

Binary outcomes → binomial(link = "logit") is the default for a reason:

**probabilities constrained to 0–1**

**coefficients interpretable as log-odds**

Other common ecological combinations:

**Counts → poisson(link = "log")**

**Overdispersed counts → negative binomial**

**Positive continuous → Gamma or lognormal**

## 4.1Visual intuition: distributions and links

```{r}
par(mfrow = c(2, 3))

x <- 0:20
plot(x, dpois(x, lambda = 5), type = "h", main = "Poisson", ylab = "Density")

plot(x, dnbinom(x, size = 2, mu = 5), type = "h", main = "Neg Binomial")

p <- seq(0.01, 0.99, length.out = 100)
plot(p, qlogis(p), type = "l", main = "Logit link", xlab = "Probability", ylab = "Linear predictor")

plot(p, log(p), type = "l", main = "Log link")

hist(rnorm(10000), breaks = 50, main = "Gaussian")

hist(rlnorm(10000), breaks = 50, main = "Lognormal")


```

*Reflection timeeeee*

Why is binomial–logit appropriate here?

Your response:

# Random effects: what they are and why we need them

A random effect is how we model structured dependence in data. Formally,
a random effect says:

> Observations within the same group are more similar to each other than
> to observations in other groups.

Instead of estimating a separate coefficient for each arena, we assume:

**arenas are drawn from a larger population of possible arenas**

**each arena has its own deviation from the global mean**

**those deviations follow a distribution (usually Normal)**

This lets us:

**model correlation correctly**

**generalize beyond the specific arenas we used**

**avoid estimating dozens of nuisance parameters**

When do we let intercepts vary?

You include a random intercept when:

**groups differ in baseline response**

**you don’t care about specific group identities**

**you want partial pooling**

Ecological examples:

**arenas**

**sites**

**individuals**

**years**

## 5.2 Modeling this in glmmTMB

For our data, a random intercept for arena looks like:

```{r}
#(1 | arena)

```

This says:

> Each arena has its own baseline probability of ticks moving toward the
> repellent, centered around a global mean.

Example for this dataset-

We expect:

**some arenas to bias ticks toward the repellent zone**

**but we don’t care which arenas those are**

That makes arena a textbook random intercept.

## 5.3 Troy is on another tangent

You might come across a "5-level rule" which is sometimes treated like a
golden rule for random effects in GLMMs. It is a rule-of-thumb that says
you need 5 or more levels to a random effect to estimate its variance
reliably. You might have convergence issues if you ignore this rule, in
ecology we ignore it all the time so take that for what it is...

# Model fitting

Before fitting, we clean up factor levels and center temperature. What
is centering? We center data by subtracting the mean from our values
which gives us a relative idea of which values are warmer or cooler than
normal. We do this in mixed-effects models specifically because a random
intercept model assumes our expected response at a value of 0 for our
predictors (or some reference value for categorical variables). Our
temperature mean is NOT 0, so we need to center it before running a
random intercept mixed-effects model. Average temperature will then = 0
but our units stay in degrees C. This is appropriate if our continuous
variables are on the same scale. What if we included Relative Humidity?
well relative humidity is bound between 0 and 100 unlike temperature and
values might be more variable than temperature. This would mean that,
regardless of how much relative humidity actually impacts our response,
it would look like it does more work than temperature just because it
has bigger numbers and varies more. Our models like big numbers and yes
they may very well be lying! Enter: scaling! We scale a variable by
dividing it by standard deviation (scale()) which gets rid of our units
entirely. This makes it much easier to compare effect sizes between
predictors at the cost of giving us an extra step to interpret. So, to
wrap up:

**Center if you have only one continuous variable in a GLMM, mandatory
if you're running a random intercept GLMM**

**Scale if you have 2 or more continuous variables and their measurement
units differ**

**Centering = subtracting mean from each value
(temp_centered=temp-mean_temp)**

**Scaling = dividing each value by a metric like standard deviation
(temp_scale=temp/temp_SD)**

**Centering means we can still interpret our model on the original scale
(change in response per 1 degree C)**

**Scaling means we now have to interpret on the standard deviation scale
or transform (change per 1 SD)**

Before we fit we will need to check what our treatment and arena
variables are labeled as and relabel them as "factors" in order to have
lme4 and glmmTMB treat the levels as comparable units. Write code to: 1)
Check what form the treatment and arena variables are currently set as
2) Convert `treatment` to a factor with `Control` as the reference
level\
3) Convert `arena` to a factor\
4) Print the levels of both

Extra challenge: try to do this with piping!

```{r}
#Insert your code here
```

```{r}
tick_mod_dat <- tick_dat %>% #just in case you got lost somewhere with the factor coding
  mutate(
    treatment = factor(treatment,
                       levels = c("Control","RepA","RepB","RepC","RepD")),
    arena = factor(arena),
    temp_c_center = temp_c - mean(temp_c)
  )
```

*NOW YOU TRY: centering vs scaling temperature (don’t just accept the
magic)*

*Write code to:* 1) Create `temp_c_center` (mean-centered)\
2) Create `temp_c_scaled` (z-scored / standardized)\
3) Print the mean and sd of both vectors\
4) Make a quick histogram of each (two plots is fine)

```{r}
#your code here
```

## 6.1 GLM (ignores grouping)

First, let’s pretend arenas don’t exist.

```{r}
m_glm <- glm(
  toward ~ treatment + temp_c_center,
  family = binomial(link = "logit"),
  data = tick_mod_dat
)

summary(m_glm)

```

*Student code space (why independence is a lie here)*

Write code to show: 1) How many ticks per trial\
2) How many trials per arena\
3) A quick table of counts by arena × treatment

```{r}
# YOUR CODE HERE
```

*Reflection*

Why might this model underestimate uncertainty?

Your response:

## 6.2 GLMM (arena + trial effects), using lme4 because variety is the spice of life

```{r}
# Random INTERCEPTS model:
# Arena and trial get their own baseline (intercept) shifts.
m_glmm <- glmer(
  toward ~ treatment + temp_c_center +
    (1 | arena) + (1 | trial_id),
  family = binomial(link = "logit"),
  data = tick_mod_dat
)

summary(m_glmm)

```

```{r}
# Random SLOPES model (by arena):
# Arenas are allowed to differ not only in baseline, but also in how strongly temperature matters.
# This is (1 + temp | arena) = correlated random intercept + random slope for temperature.
m_glmm_rs <- glmer(
  toward ~ treatment + temp_c_center +
    (1 + temp_c_center | arena) + (1 | trial_id),
  family = binomial(link = "logit"),
  data = tick_mod_dat
)

summary(m_glmm_rs)

```

Quick checks + comparison (optional but useful)

```{r}
# 1) Convergence / singularity checks
isSingular(m_glmm)
isSingular(m_glmm_rs)
# 2) Look at variance components:
# - Random intercept variance: do arenas differ in baseline?
# - Random slope variance: do arenas differ in temp sensitivity?
# - Correlation: do high-baseline arenas also have stronger/weaker temp effects?
VarCorr(m_glmm)
VarCorr(m_glmm_rs)
# 3) Compare models:
# LRT is valid here because random-intercepts model is nested within random-slopes model.
anova(m_glmm, m_glmm_rs, test = "Chisq")

# Also show AIC as a quick check
AIC(m_glmm, m_glmm_rs)
```

# Convergence: what it is and why you should care

What does convergence mean? Convergence is a model state that is reached
when the numerical optimizer inside your GLMM function has found the
most likely parameters to give the data points you have. In
GLMM/regression world the optimization algorithm used is called Maximum
Likelihood Estimation (MLE), that is basically all I know plus that the
glmmTMB optimizer tends to converge faster than lme4 and that one of the
primary differences in how the various glmm packages operate is through
their different optimization algorithms to impliment MLE.

A model converges when the optimizer believes it has found parameter
values that:

**maximize the likelihood**

**are stable**

**satisfy numerical tolerances**

Convergence is not “this model is correct” — it’s:

> “The math didn’t obviously fail.”

What non-convergence usually means

**model too complex for the data**

**random effect variance near zero**

**separation in binomial data**

**collinearity**

If your model converges, that does not mean it is all good to go! Still
need to check fit, dispersion, etc. but a lack of convergence is a clear
sign of issues. \# 7.1 Singularity

```{r}
isSingular(m_glmm)

```

Write code to: 1) Check singularity for both models (random intercepts
vs random slopes if you fit them)\
2) Extract `VarCorr()` and identify which variance component is \~0\
3) Print the random effects for arena (`ranef`) and interpret which
arena is “highest”

```{r}
# YOUR CODE HERE
```

#7.2 Random-effect variance

```{r}
VarCorr(m_glmm)
ranef(m_glmm)$arena
```

#Interpretation

Is arena variance meaningful?

Your response:

#7.3 Residual diagnostics Simulation-based residuals help us evaluate:

**dispersion**

**zero inflation**

**systematic patterns**

```{r}
library(DHARMa)
res <- simulateResiduals(m_glmm)
plot(res)
```

Reflection

What kinds of problems can simulation-based residuals detect?

Your response:

# Interpreting repellent effects

*Interpretation of the Tick Movement Logistic Model*

We fit a logistic regression model predicting whether a tick moved
toward the repellent zone (1) versus away (0) as a function of treatment
and centered temperature:

$logit(p)=β0+βtreatment+β1tempcenter$

*Intercept*

Estimate: 0.426

The intercept represents the expected log-odds that a tick moves toward
the repellent when:

Treatment = Control (reference level)

Temperature = average temperature (because we centered it)

Back-transforming to probability:

$p=e0.4261+e0.426≈0.60$

So, under control conditions at average temperature, the predicted
probability that a tick moves toward the repellent zone is approximately
60%.

Note: The intercept is not statistically significant (p = 0.17), but
intercept significance is rarely biologically meaningful in this
context, it is primarily a baseline reference point.

*Treatment Effects*

Each treatment coefficient represents the change in log-odds relative to
the control, holding temperature constant.

*RepA*

$Estimate: −0.764$ $Odds ratio: exp(−0.764) ≈ 0.47$

We are exponentiating our estimates to convert them to a probability
percentage (easier to talk about) Interpretation: Ticks exposed to RepA
have approximately 53% lower odds of moving toward the repellent zone
compared to control, at the same temperature.

This effect is statistically significant ($p = 0.041$).

*RepB*

$Estimate: −1.794$ $Odds ratio: exp(−1.794) ≈ 0.17$

Interpretation: Ticks exposed to RepB have about 83% lower odds of
moving toward the repellent zone compared to control, holding
temperature constant.

This is a strong and highly significant effect ($p < 0.001$).

*RepC*

$Estimate: −0.572$ $Odds ratio: exp(−0.572) ≈ 0.56$

Interpretation: RepC is associated with a 44% reduction in odds of
moving toward the repellent zone compared to control, but this effect is
not statistically significant ($p = 0.14$).

*RepD*

$Estimate: −1.381$ $Odds ratio: exp(−1.381) ≈ 0.25$

Interpretation: RepD reduces the odds of moving toward the repellent
zone by approximately 75% relative to control, holding temperature
constant. This effect is statistically significant (p \< 0.001).

*Temperature Effect*

$Estimate: 0.218$

For each 1°C increase above the average temperature:

The log-odds of moving toward the repellent increase by 0.218

$Odds ratio: exp(0.218) ≈ 1.24$

Interpretation: For each 1°C increase above the average temperature, the
odds that a tick moves toward the repellent zone increase by
approximately 24%, holding treatment constant.

This effect is highly significant (p \< 0.001).

Big Picture Interpretation

Across treatments, several repellents significantly reduce the
probability that ticks move toward the treated zone, with RepB and RepD
showing the strongest effects.

However, temperature independently increases the probability that ticks
move toward the repellent area, suggesting that behavioral responses are
not purely treatment-driven but are also influenced by environmental
conditions.

Because this is a logistic model, coefficients are interpreted on the
log-odds scale. As with other GLMs, we can interpret them directly in
log-odds, or exponentiate them to obtain odds ratios, which are often
easier to explain. \# 8.1 Odds ratios

```{r}
fixef_tab <- tidy(m_glmm, effects = "fixed", conf.int = TRUE) %>%
  mutate(
    odds_ratio = exp(estimate),
    lo = exp(conf.low),
    hi = exp(conf.high)
  )

fixef_tab
```

*Student code space (odds ratios → words)*

Write code to: 1) Pull the odds ratio for `RepB` (or whichever repellent
you choose)\
2) Convert it to a percent change in odds vs control: `(OR - 1) * 100`\
3) Write one sentence interpretation *in a comment* next to your code

```{r}
# YOUR CODE HERE
```

# 8.2 Predicted probabilities

Odds ratios are useful, but probabilities are often more intuitive.

```{r}
newdat <- expand_grid(
  treatment = levels(tick_mod_dat$treatment),
  temp_c_center = 0
) %>%
  mutate(
    logit = predict(m_glmm, newdata = ., type = "link", re.form = NA),
    prob  = plogis(logit)
  )

newdat

newdat %>%
  ggplot(aes(treatment, prob)) +
  geom_point(size = 3) +
  ylim(0, 1) +
  theme_bw()
```

Often times you'll see odds-ratios reported in stats sections, but the
pros will often go one-step further and interpret probabilities using
the predict function above.

*Interpretation*

What is the absolute probability difference between control and the
strongest repellent?

Your response:

# 9. Final takeaway

GLMMs combine likelihood choice + random effects

Random effects model dependence, not nuisance

Interpretation should focus on probabilities and comparisons

Model checking is non-optional

# Final reflection

What would you change first if diagnostics looked poor?

Your response:

Literature/Resources:

-   Actual vignette for fitting glmms in glmmTMB:
    <https://cran.r-project.org/web/packages/glmmTMB/glmmTMB.pdf> -Very
    good description of error distributions:
    <https://r.qcbs.ca/workshop07/book-en/choose-an-error-distribution.html>
-   Slightly notation-heavy explanation of generalized linear model,
    great odds ratio interpretation explanation:
    <https://stats.oarc.ucla.edu/other/mult-pkg/introduction-to-generalized-linear-mixed-models/>
-   Regression and other Stories Chapters 6-8
    <https://users.aalto.fi/~ave/ROS.pdf>
-   Very good explanation of mixed models and hierarchy/grouping/slopes
    and intercepts:
    <https://ourcodingclub.github.io/tutorials/mixed-models/>
